\chapter*{Preface}

\markboth{PREFACE}{PREFACE}
\addcontentsline{toc}{chapter}{Preface}

\begin{quote}
This book is a work in progress. This {\bf Preface} will be the last chapter written. Please skip to {\bf Chapter 1} to see some progress.  -- LR
\end{quote}

{\it Think Go} is an introduction to computer science and programming intended for readers with little or no experience.
We start with the most basic concepts and are careful to define all terms when they are first used.
The book presents each new idea in a logical progression.
Larger topics, like control flow statements and interfaces, are divided into smaller examples and introduced over the course of several chapters.

This book is intentionally concise.
Each chapter is 12--14 pages and covers the material for one week of a college course.
It is not meant to be a comprehensive presentation of Go, but rather, an initial exposure to programming constructs and techniques.
We begin with small problems and basic algorithms and work up to program and library design.


\section*{The philosophy behind the book}

The chapters of this book are meant to be read in order, because each one builds on the previous one.

Here are the guiding principles that make the book the way it is:

\begin{itemize}

\item {\em One concept at a time.}
We break down topics that give beginners trouble into a series of small steps, so that they can exercise each new concept in isolation before continuing.

\item {\em Balance of Go and concepts.}
The book is not primarily about Go; it uses code examples to demonstrate computer science.
Most chapters start with language features and end with concepts.

\item {\em Conciseness.}
An important goal of the book is to be small enough so that students can read and understand the entire text in a one-semester college or AP course.
%Students can read 1--2 chapters per week, depending on the pace of the instruction.

\item {\em Emphasis on vocabulary.}
We try to introduce the minimum number of terms and define them carefully when they are first used.
We also organize them in glossaries at the end of each chapter.

\item {\em Program development.}
There are many strategies for writing programs, including bottom-up, top-down, and others.
We demonstrate multiple program development techniques, allowing readers to choose methods that work best for them.

\item {\em Multiple learning curves.}
To write a program, you have to understand the algorithm, know the programming language, and be able to debug errors.
We discuss these and other aspects throughout the book, and include an appendix that summarizes our advice.

\item {\em Spiral approach.}
Some concepts take time to sink in.
The more difficult ideas in the book, like recursion, appear several times.
By coming back to these topics, we give learners a chance to review and reinforce.

%\item {\em Keep it simple.}
%We use the minimum amount of Java to get the maximum amount of programming ability.
%The goal of this book is to teach fundamental ideas from computer science, not Java.
%We leave out some language features, like the \java{switch} statement, that are unnecessary, and we present only a few of the many classes in the Java libraries.

\end{itemize}

\section*{About the appendixes}

We also include several appendixes with material that can be read at any time:

\begin{description}

\item{\bf Appendix D: Debugging}

We provide debugging suggestions throughout the book, but we also have an appendix with many more suggestions on how to debug your programs.
We recommend that readers review this appendix frequently as they work through the book.

\end{description}


\section*{Using the code examples}
\label{code}

Everything after this paragraph is unedited text from {\it Think Java 2e}. Please forgive the noise and dust while we remix that book and {\it Think Python 2e} to make {\it Think Go}.

Most of the code examples in this book are available from a Git repository at \url{https://github.com/ChrisMayfield/ThinkJavaCode2}.
Git is a ``version control system'' that allows you to keep track of the files that make up a project.
A collection of files under Git's control is called a ``repository''.

\index{repository}
\index{GitHub}

GitHub is a hosting service that provides storage for Git repositories and a convenient web interface.
It provides several ways to work with the code:

\begin{itemize}

\item You can create a copy of the repository on GitHub by pressing the {\sf Fork} button.
If you don't already have a GitHub account, you'll need to create one.
After forking, you'll have your own repository on GitHub that you can use to keep track of code you write.
Then you can ``clone'' the repository, which downloads a copy of the files to your computer.

\item Alternatively, you could clone the original repository without forking.
If you choose this option, you don't need a GitHub account, but you won't be able to save your changes on GitHub.

\item If you don't want to use Git at all, you can download the code in a ZIP archive using the {\sf Download ZIP} button on the GitHub page, or this link: \url{http://tinyurl.com/ThinkJavaCode2}.

\end{itemize}

After you clone the repository or unzip the ZIP file, you should have a directory named {\tt ThinkJavaCode2} with a subdirectory for each chapter in the book.

All examples in this book were developed and tested using Java SE Development Kit 8.
If you are using a more recent version, the examples in this book should still work.
If you are using an older version, some of them may not.


\section*{Contributors over the years}

Many people have sent corrections and suggestions, and we appreciate their valuable feedback!
This list begins with Version 4.0 of the open-source edition, so it omits those who contributed to earlier versions.

\begin{itemize}

\item Ellen Hildreth used this book to teach Data Structures at Wellesley College and submitted a whole stack of corrections, along with some great suggestions.

\item Tania Passfield pointed out that some glossaries had leftover terms that no longer appeared in the text.

\item Elizabeth Wiethoff noticed that the series expansion of $\exp(-x^2)$ was wrong.
She has also worked on a Ruby version of the book.

\item Matt Crawford sent in a whole patch file full of corrections.

\item Chi-Yu Li pointed out a typo and an error in one of the code examples.

\item Doan Thanh Nam corrected an example.

\item Muhammad Saied translated the book into Arabic, and found several errors in the process.

\item Marius Margowski found an inconsistency in a code example.

\item Leslie Klein discovered another error in the series expansion of $\exp(-x^2)$, identified typos in the card array figures, and gave helpful suggestions to clarify several exercises.

\item Micah Lindstrom reported half a dozen typos and sent corrections.

\item James Riely ported the textbook source from LaTeX to Sphinx.
\\ \url{http://fpl.cs.depaul.edu/jriely/thinkapjava/}

\item Peter Knaggs ported the book to C\#.
\\ \url{http://www.rigwit.co.uk/think/sharp/}

\item Heidi Gentry-Kolen recorded several video lectures that follow the book.
\\ \url{https://www.youtube.com/user/digipipeline}

\item Waldo Ribeiro submitted a pull request that corrected a dozen typos.
\end{itemize}

We are especially grateful to the technical reviewers of the O'Reilly Media first edition: Blythe Samuels, David Wisneski, and Stephen Rose.
They found errors, made many great suggestions, and helped make the book much better.

Many students have given exceptional feedback, including Ian Staton, Tanner Wernecke, Jacob Green, Rasha Abuhantash, Nick Duncan, Kylie Davidson, Shirley Jiang, Elena Trafton, Jennifer Gregorio, and Azeem Mufti.

Other contributors who found one or more typos: Stijn Debrouwere, Guy Driesen, Andai Velican, Chris Kuszmaul, Daniel Kurikesu, Josh Donath, Rens Findhammer, Elisa Abedrapo, Yousef BaAfif, Bruce Hill, Matt Underwood, Isaac Sultan, Dan Rice, Robert Beard, Daniel Pierce, Michael Giftthaler, Chris Fox, Min Zeng, and Markus Geuss.

%\medskip

If you have additional comments or ideas about the text, please send them to: \href{mailto:feedback@greenteapress.com}{\tt feedback@greenteapress.com}.

Allen Downey and Chris Mayfield
